Detecting Stylistic Variation in Pop Production using the Spotify API

### 2023-11-08 outputs/paper/paper.qmd
paper edit
edits to paper

---
title: "title"
subtitle: "subtitle"
author: 
  - Michaela Drouillard
thanks: "Code and data are available at: https://github.com/michaeladrouillard/spotify_pop."
date: today
abstract: ""
format: pdf
number-sections: true
bibliography: references.bib
---

@@

@@

I spoke to Gamman to better undertstand makes a song Antonoff-ish, and to try to understand his reaction to it. In trying to understand how to describe Antonoff's sound, I used the Spotify API's audio features dataset to explore difference machine listening metrics, and build a model to recreate a digital version of Gamman's classification excercise. I also spoke with Glen McDonald, an engineer at Spotify, from whom I learned that the `valence` and `danceability` variables in the Spotify API were created by hiring college interns and having them tag whether songs were positive or danceable.

I used two models to investigate the data and ask the question: What is Antonoff's sound? Can we pin down style using the quantitative metrics offered by Spotify?

The first model, a logistic regression, draws on the dataset of artists who've collaborated with Antonoff, with an extra binary variable to tag whether or not the song was an Antonoff production. With this model, we can get a first glimpse at variables that are stronger predictors of Antonoff-iness, along with a better understanding of whether the Spotify data can trace Antonoff's influence on an artist's sound over time.

@@

I built a second model, a Random Forest, to complement the insights drawn from the logistic regression. The Random Forest trains on the full data set, including other major pop producers. Random Forests don't assume linear relationships, and aren't sensitive to multicollinearity. In fact, using the tk package in R, we can visualize and analyse the relationships between different variables. The variance in correlation scores across producers can be taken as an approximation for style, since this variance helps the model distinguish between producers.

In the discussion section, I listen to the songs with the highest danceability scores, and make an argument that this `danceability` feature, which has so much uncertainty cooked into it, is really capturing something like visceral-ness. Against the odds, we can capture something about Antonoff's music with these variables, and it might, in the end, just be that subjective "ick" that some people feel, and others enjoy, and that we're all just trying to describe in using different methods and tools.

# tk: Lit Review

@@

The audio feature data was pulled using Spotify API, which I accessed using the `spotifyr` package in R [@citeSpotify]. I used the `get_artist_audio_features` function to acquire data on Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, and Bleachers discographies. I threw in HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Von Etten, and Mitski in too, since they've never collaborated with Jack Antonoff but are more or less part of the same pop sphere.

@@

### Artist Selection

@@

### Data Cleaning

I joined these data sets using the `rbind` function in base R, and created an additional variable called "jack", which contains a 1 if Jack Antonoff produced or co-produced the song, and a 0 if he had nothing to do with it. I got the information on which songs Antonoff did or did not produce from the "Jack Antonoff Production Discography" Wikipedia page [@citewiki].

@@

## Interview with Glen McDonald

@@

# Logistic Regression

## Data Preparation

@@

@fig-means gives us an idea of the mean value of each of these features for Antonofff and non-Antonoff-produced tracks.

@@

## Results

@@

# Random Forest

Random forests (RF) build an ensemble of decision trees during the training phase, and then use majority voting for classification tasks. Each tree in the ensemble is trained on a random sample of data taken with replacement, and only a random subset of features are considered when splitting at each node. This method ensures that each tree in the ensemble is unique, avoiding overfitting and capturing non-linear relationships and more complex patterns in the data. The formula behind the RF is:

## Data Preparation

## Model Training

## Results

### Model Performance

@@

### Stylistic Difference

>>>>

---
title: "Detecting Stylistic Variation in Pop Production using the Spotify API"
author: 
  - Michaela Drouillard
thanks: "Code and data are available at: https://github.com/michaeladrouillard/spotify_pop."
date: today
abstract: "This paper explores whether or not we can predict whether a song was produced by Jack Antonoff using the audio features provided by Spotify's API. We find that in general our model has an accuracy score of 90%. (Precision and recall) We interestingly found that danceability, a metric trained on human-tagged data, was a key predictor. Our results have implications for the understanding of culture, how the combination human subjectivity and machine listening can come to create an approximate for style."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

We're interested in understanding how to create an approximate for production style. To do this we used Spotify API's audio features dataset to explore different audio feature metrics and identify which were the most import predictors in identifying pop producers' style. We chose Jack Antonoff as focus, as he is a prolific and widely used producer, who also has his own solo acts.

By way of background, Antonoff began his career as a guitarist in fun., and is the frontman in Bleachers. He has produced and co-produced music for Lorde, Taylor Swift, Florence and the Machine, The Chicks, Clairo, The 1975, Grimes, Zayn, Pink, Lana Del Rey, Olivia Rodrigo, the Minions: The Rise of Gru Soundtrack, and more. His style is considered identifiable. As Andrew Marantz writes in the *New Yorker*: "When his band releases an album, the world responds politely. When he produces one by Lorde or Lana Del Rey or Taylor Swift, the world wobbles on its axis" [@citeMarantz].

We are interested in whether Spotify audio features can be used to train a model that distinguishes whether a song was produced by Antonoff. We obtained tracks from six artists who've collaborated with Antonoff, both before their collaboration and during. We also obtained tracks from similarly prolific pop producers. The producers we compare Antonoff's production style with are Little, Rechstad, Martin, Kurstin, Epworth and Nowels. Our data set contains 3738 tracks in total.

We built two models to explore the data. The first model, a logistic regression, draws on the dataset of artists who've collaborated with Antonoff, with a binary variable to tag whether or not the song was an Antonoff production. With this model, we explore which variables are stronger predictors of Antonoff's style, developping a better understanding of whether the Spotify data can trace Antonoff's influence on an artist's sound over time. We built a second model, a Random Forest (RF), to complement the insights drawn from the logistic regression. The RF trains on the full data set, including other major pop producers.

We find that danceability is the strongest predictor of whether or not a song has been produced by Antonoff. Furthermore, when the variance is considered along with other features of the dataset, the variance tends to be strongest classifier of whether a song was produced by Antonoff. Daanceability in concert with other features in the dataset provides the best predictions.

Against the odds, we can capture something about Antonoff's music with these variables, and it might, in the end, just be that subjective "ick" that some people feel, and others enjoy, and that we're all just trying to describe in using different methods and tools. The remainder of this paper is structured as follows: @ref-background...

# Background {#sec-background}

@@

I spoke to Gamman to better undertstand makes a song Antonoff-ish, and to try to understand his reaction to it.

I also spoke with Glen McDonald, an engineer at Spotify, from whom I learned that the `valence` and `danceability` variables in the Spotify API were created by hiring college interns and having them tag whether songs were positive or danceable.

@@

In the discussion section, I listen to the songs with the highest danceability scores, and make an argument that this `danceability` feature, which has so much uncertainty cooked into it, is really capturing something like visceral-ness.

@@

The audio feature data was pulled using Spotify API, which I accessed using the `spotifyr` package in R [@citeSpotify]. I used the `get_artist_audio_features` function to acquire data on Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, and Bleachers discographies. HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Von Etten, and Mitski were included, since they've never collaborated with Jack Antonoff but are more or less part of the same pop sphere.

@@

@@

We created a binary variable which contains 1 if Jack Antonoff produced or co-produced the song, and a 0 if it was produced by someone else. The information underpinning this discography was drawn from the "Jack Antonoff Production Discography" Wikipedia page [@citewiki].

@@

In total 3738 tracks from X different artists. (Producers, etc)

@@

# Model

## Logistic Regression

@@

@@

## Random Forest

Random forests (RF) build an ensemble of decision trees during the training phase, and then use majority voting for classification tasks. Each tree in the ensemble is trained on a random sample of data taken with replacement, and only a random subset of features are considered when splitting at each node. This method ensures that each tree in the ensemble is unique, avoiding overfitting and capturing non-linear relationships and more complex patterns in the data.

add model here

# Results

## Logistic Regression

@@

## Random Forest

@@
